---
title: ' Community ecology - Computer lab II - AB332'
author: "Ramiro Logares (ICM), with additions by Anders K. Krabberød (UiO)"
date: "October 2021"
output: html_notebook
---
# Computer lab II

First load the necessary packages.
```{r,include=FALSE}
library(vegan)
library(tidyverse)
library(compositions)
library(zCompositions)
library(PCAtools)
library(mixOmics)
library(ape)
library(recluster)
library(dendextend)
```

Now load the the data from the previous lab (change path to the file if necessary):
```{r}
load("AB332_lab_I.RData")
```


# Ordination and clustering

## PCA
Pca from the rarefied table
```{r}
otu.tab.simple.ss.nozero.pca <- PCAtools::pca(t(otu.tab.simple.ss.nozero), scale = FALSE) # Runs de PCA

biplot(otu.tab.simple.ss.nozero.pca, showLoadings = T, lab = rownames(otu.tab.simple.ss.nozero)) # Plots de PCA
screeplot(otu.tab.simple.ss.nozero.pca, axisLabSize = 18, titleLabSize = 22) # We plot the percentage of variance explained by each axis
```

PCA on the clr table
```{r}
otu.tab.simple.gbm.clr.pca <- mixOmics::pca(otu.tab.simple.gbm.clr, scale = FALSE, ncomp = 6) # NB: the pca used here is from "mixOmics" while the pca above is from "PCAtools"
plotVar(otu.tab.simple.gbm.clr.pca)
plot(otu.tab.simple.gbm.clr.pca)
```

## PCoA

PCoA not included in Vegan, so we use the APE package
Calculates the PCoA of the rarefied table using Bray Curtis distances

```{r}
otu.tab.simple.ss.nozero.bray.pcoa <- pcoa(otu.tab.simple.ss.nozero.bray) # Calculates PCoA
barplot(otu.tab.simple.ss.nozero.bray.pcoa$values$Relative_eig[1:10]) # Plot Eigenvalues (amount of variance explained by Axes)
biplot.pcoa(otu.tab.simple.ss.nozero.bray.pcoa)
```

Calculates the PCoA of the clr table using Euclidean distances
```{r}
otu.tab.simple.gbm.clr.euclidean.pcoa <- pcoa(otu.tab.simple.gbm.clr.euclidean) # Calculates PCoA
barplot(otu.tab.simple.gbm.clr.euclidean.pcoa$values$Relative_eig[1:10]) # Plot Eigenvalues (amount of variance explained by Axes)
biplot.pcoa(otu.tab.simple.gbm.clr.euclidean.pcoa)
```
## NMDS
We will define the function NMDS.scree() that automatically performs a NMDS for 1-7 dimensions and plots the number of dimensions vs. stress


```{r}

set.seed(666) # Set a seed to make results reproducible
NMDS.scree <- function(x) { # x is the name of the distance matrix
  plot(rep(1, 7), replicate(7, metaMDS(x, autotransform = FALSE, k = 1)$stress), xlim = c(1, 7), ylim = c(0, 0.30), xlab = "# of Dimensions", ylab = "Stress", main = "NMDS stress plot")
  for (i in 1:7) {
    points(rep(i + 1, 7), replicate(7, metaMDS(x, autotransform = FALSE, k = i + 1)$stress))
  }
}
```

Using the function to determine the optimal number of dimensions
Using the rarefied table
```{#r,include=FALSE}
NMDS.scree(otu.tab.simple.ss.nozero.bray)
```

Using the clr table
```{#r}
NMDS.scree(otu.tab.simple.gbm.clr.euclidean)
```

We calculate NMDS for k(dimensions)=2
Rarefied table (we use the dataframe to have access to sample and OTU names)
```{r}
otu.tab.simple.ss.nozero.bray.nmds <- metaMDS(otu.tab.simple.ss.nozero, k = 2, trymax = 100, trace = FALSE, autotransform = FALSE, distance = "bray")
otu.tab.simple.ss.nozero.bray.nmds
```


```{r}
stressplot(otu.tab.simple.ss.nozero.bray.nmds) # Make stressplot
```

clr table (we use the dataframe to have access to sample and OTU names)
```{r}
otu.tab.simple.gbm.clr.euclidean.nmds <- metaMDS(t(as.data.frame(otu.tab.simple.gbm.clr)), k = 2, trymax = 100, trace = FALSE, autotransform = FALSE, distance = "euclidean")
otu.tab.simple.gbm.clr.euclidean.nmds
```


```{r}
stressplot(otu.tab.simple.gbm.clr.euclidean.nmds) # Make stressplot
```

Simple plotting
Rarefied table
```{r}
plot(otu.tab.simple.ss.nozero.bray.nmds, display = "sites", type = "n")
points(otu.tab.simple.ss.nozero.bray.nmds, display = "sites", col = "red", pch = 19)
text(otu.tab.simple.ss.nozero.bray.nmds, display = "sites")
```

clr table
```{r}

plot(otu.tab.simple.gbm.clr.euclidean.nmds, display = "sites", type = "n")
points(otu.tab.simple.gbm.clr.euclidean.nmds, display = "sites", col = "red", pch = 19)
text(otu.tab.simple.gbm.clr.euclidean.nmds, display = "sites")
```
### Let's make nicer plots
We get the seasons for samples

```{r}
isa.metadata <- read_tsv("https://raw.githubusercontent.com/krabberod/UNIS_AB332_2021/main/computer_lab/data/AB332metadata_v3.txt")
isa.metadata <- column_to_rownames(isa.metadata, var = "Sample_Name")
isa.metadata.simp <- isa.metadata[6:30, ]
```

Rarefied table
We generate a table of nmds scores and other features

```{r}
otu.tab.simple.ss.nozero.bray.nmds.scores <- as.data.frame(scores(otu.tab.simple.ss.nozero.bray.nmds))
otu.tab.simple.ss.nozero.bray.nmds.scores$season <- isa.metadata.simp$seasons
otu.tab.simple.ss.nozero.bray.nmds.scores$month <- as.factor(isa.metadata.simp$month)
otu.tab.simple.ss.nozero.bray.nmds.scores$samples <- rownames(otu.tab.simple.ss.nozero.bray.nmds.scores)
```

Create the plot
```{r}
ggplot(otu.tab.simple.ss.nozero.bray.nmds.scores) +
  geom_point(mapping = aes(x = NMDS1, y = NMDS2, colour = season), size = 3) +
  coord_fixed() + ## need aspect ratio of 1!
  geom_text_repel(
    box.padding = 0.5, aes(x = NMDS1, y = NMDS2, label = samples),
    size = 3
  )
```

clr table
We generate a table of nmds scores and other features
```{r}
otu.tab.simple.gbm.clr.euclidean.nmds.scores <- as.data.frame(scores(otu.tab.simple.gbm.clr.euclidean.nmds))
otu.tab.simple.gbm.clr.euclidean.nmds.scores$season <- isa.metadata.simp$seasons
otu.tab.simple.gbm.clr.euclidean.nmds.scores$month <- as.factor(isa.metadata.simp$month)
otu.tab.simple.gbm.clr.euclidean.nmds.scores$samples <- rownames(otu.tab.simple.gbm.clr.euclidean.nmds.scores)
otu.tab.simple.gbm.clr.euclidean.nmds.scores
```


```{r}
ggplot(otu.tab.simple.gbm.clr.euclidean.nmds.scores) +
  geom_point(mapping = aes(x = NMDS1, y = NMDS2, colour = month), size = 3) +
  coord_fixed() + ## need aspect ratio of 1!
  geom_text_repel(
    box.padding = 0.5, aes(x = NMDS1, y = NMDS2, label = samples),
    size = 3
  )
```


#Clustering of samples

Allows determining the similarity between samples as well as the organization of samples in groups.
Hierarchical clustering: samples will be organized in ranks according to their similarity and all samples will be included in a large group Unweighted Pair-Group Method Using Arithmetic Averages (UPGMA): This linkage method will link samples by considering their distance to a subgroup arithmetic average. This is a method widely used in ecology

### UPGMA
Rarefied dataset
We generate 100 trees by re-sampling and then, we plot the consensus tree
```{r}
otu.tab.simple.ss.nozero.bray.upgma <- recluster.cons(otu.tab.simple.ss.nozero.bray, tr = 100, p = 0.5, method = "average")
plot(otu.tab.simple.ss.nozero.bray.upgma$cons) # plot consensus tree
```

We'll calculate bootstrap support values (0: bad - 100: perfect)
This allows us to know how well supported is the branching pattern
```{r}
otu.tab.simple.ss.nozero.bray.upgma.boot <- recluster.boot(otu.tab.simple.ss.nozero.bray.upgma$cons, otu.tab.simple.ss.nozero, tr = 100, p = 0.5, method = "average", boot = 1000, level = 1)
recluster.plot(otu.tab.simple.ss.nozero.bray.upgma$cons, otu.tab.simple.ss.nozero.bray.upgma.boot) # We add bootstrap values to the branching pattern
```
clr transformed dataset
We generate 100 trees by resampling and then, we build the consensus

```{r}
otu.tab.simple.gbm.clr.euclidean.upgma <- recluster.cons(otu.tab.simple.gbm.clr.euclidean, tr = 100, p = 0.5, method = "average")
plot(otu.tab.simple.gbm.clr.euclidean.upgma$cons) # plot consensus tree
```
 We'll calculate bootstrap support values (0: bad - 100: perfect)

```{r}
otu.tab.simple.gbm.clr.euclidean.upgma.boot <- recluster.boot(otu.tab.simple.gbm.clr.euclidean.upgma$cons, t(otu.tab.simple.gbm.clr), tr = 100, p = 0.5, method = "average", boot = 100, level = 1)

recluster.plot(otu.tab.simple.gbm.clr.euclidean.upgma$cons, otu.tab.simple.gbm.clr.euclidean.upgma.boot) # We add bootstrap values to the branching pattern
```
Let's compare both dendrograms using tanglegrams


```{r}

dendlist(as.dendrogram(otu.tab.simple.ss.nozero.bray.upgma$cons), as.dendrogram(otu.tab.simple.gbm.clr.euclidean.upgma$cons)) %>%
  untangle(method = "step1side") %>% # Find the best alignment layout
  tanglegram(cex_main = 0.7, cex_sub = 1, lwd = 2.0, main_left = "rarefied", main_right = "clr transformed", cex_main_left = 2, lab.cex = 1.5, edge.lwd = 2)
```



```{r}
save.image("AB332_lab_II.RData")
```
